Implications of afforestation policies for global markets

### 2023-06-24 paper.Rmd
Update paper.Rmd

In 2020, there were 4.1 billion hectares of forestland globally that cover roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink, currently sequestering about 30% of GHG emmissions [@roe2019]. With increasingly negative climate impacts, the pressure to enhance the terrestrial sink is growing, and one of the tools to achieve this is afforestation. There is growing attention being placed globally to afforestation efforts, which is highligted by the UN's declaration of the years from 2021 through 2030 as the UN Decade of Ecosystem Resotration [@sewell2020], and many of the Nationally Determined Contributions (NDCs) countries have submitted, which detail their plans to keep warming under the 2C threshold, rely to a significant degree on afforestation projects [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021].

The global gains that can be realized from afforestion, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@doelman2020].  In many cases, these results suggest massive losses in pasture and cropland. This puts into question the feasibility of attaining these results, as forests are continually threatened by issues such as rising incomes and food prices and population growth that put pressure on policymakers to allow greater conversion of forestland into farmland. Previous research though has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017]. Still, the manner in which these various and oftentimes competing concerns are addressed will have significant impacts on the global community.

Moreover, afforestation appears to not provide the same benefits in all regions of the world. In particular, boreal forests appear to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017] [@li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. With tropical forests, a strong positive result is derived from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its committments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its masssive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019].

>>>>

In 2020, there were 4.1 billion hectares of forestland globally that cover roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink, currently sequestering about 30% of GHG emissions [@roe2019]. With increasingly negative climate impacts, the pressure to enhance the terrestrial sink is growing, and one of the tools to achieve this is afforestation. There is growing attention being placed globally to afforestation efforts, which is highlighted by the UN's declaration of the years from 2021 through 2030 as the UN Decade of Ecosystem Restoration [@sewell2020], and many of the Nationally Determined Contributions (NDCs) countries have submitted, which detail their plans to keep warming under the 2C threshold, rely to a significant degree on afforestation projects [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021].

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@doelman2020].  In many cases, these results suggest massive losses in pasture and cropland. This puts into question the feasibility of attaining these results, as forests are continually threatened by issues such as rising incomes and food prices and population growth that put pressure on policymakers to allow greater conversion of forestland into farmland. Previous research though has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017]. Still, the manner in which these various and oftentimes competing concerns are addressed will have significant impacts on the global community.

Moreover, afforestation appears not to provide the same benefits in all regions of the world. In particular, boreal forests appear to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017] [@li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. With tropical forests, a strong positive result is derived from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its commitments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its massive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019].
